---
title: "STT465Hw4"
author: "Sam Isken"
date: "October 27, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

STT 465 Fall 2019
Homework 4
Due 10/30/2019 (In Class)

Instruction:
When using R in any problem, copy the code and results onto your word document under that question number and add any required comments. You will lose points if I do not see your codes.

You should present a stapled document when multiple pages are used. The grader will not be held responsible for any loss of pages.

The gout data set (in D2L) contains information on the gout status of individual with respect to their sex, race, age and serum urate (su).
After downloading onto your computer, import the data into R using:

```{r}
GOUT <- read.table("gout.txt",header=TRUE)
## Transform qualitative variables into factors in R, for example:
GOUT$sex <- factor(GOUT$sex,levels=c('M','F'))
pairs(GOUT)
summary(GOUT)
head(GOUT)
```

1. Using the gout data set estimate the effects of sex, race and age on serum urate using the lm() function.

```{r}
Model1_Gout <- lm(su~sex+race+age, data=GOUT)
Model1_Gout
```

-Given a 1 unit (year) increase in age there is predicted to be a .02674 increase in su. 
-Given the sex of a patient is female there is predicted to be a 1.52853 DECREASE (-1.52853) in su. 
-Given the race of a patient is white there is predicted to be a .78212 DECREASE (-.78212) in su. 

(a) Report your results.

```{r}
Model1_Gout
summary(Model1_Gout)
coefficients(Model1_Gout)
```

The initial model is: 

$$ \text{su}=\beta_0 + \text{sexF}*x_1 + \text{raceW}*x_2 + \text{age} * x_3$$
After calculating predicted values we can show our model to be: 

$$ Y = -1.52853 x_1 + -0.78212 x_2 + 0.02674 x_3 +5.84828  $$ 

(b) Summarize your findings.

-Given a 1 unit (year) increase in age there is predicted to be a .02674 increase in su. 
-Given the sex of a patient is female there is predicted to be a 1.52853 DECREASE (-1.52853) in su. 
-Given the race of a patient is white there is predicted to be a .78212 DECREASE (-.78212) in su. 


The initial model is: 

$$ \text{su}=\beta_0 + \text{sexF}*x_1 + \text{raceW}*x_2 + \text{age} * x_3$$
After calculating predicted values we can show our model to be: 

$$ Y = -1.52853 x_1 + -0.78212 x_2 + 0.02674 x_3 +5.84828  $$ 

All of the predictors are significant with age being the least significant. 

(c) Report the code you used to obtain the results in (a) and (b)

```{r}
Model1_Gout <- lm(su~sex+race+age, data=GOUT)
Model1_Gout
summary(Model1_Gout)
coefficients(Model1_Gout)
```

LaTeX Code:

```{LaTeX}
# Initial Model 
\text{su}=\beta_0 + \text{sexF}*x_1 + \text{raceW}*x_2 + \text{age} * x_3
# Filled in Model
Y = -1.52853 x_1 + -0.78212 x_2 + 0.02674 x_3 +5.84828  
```


2. Produce a table equal to summary( ) using matrix operations. Your table must include parameter estimate, SE, t-statistic and p-value.

```{r}
# Incidence matrix for intercept and effects of Sex, Race, Age,
X=cbind(1,GOUT$sex,GOUT$race, GOUT$age) 
X1 <- model.matrix(~sex+race+age,data=GOUT)
head(X)
y=GOUT$su
head(y)
dim (X)
y = as.matrix(y)
dim(y)
# OLS equations
Xy=t(X)%*%y
XtX=t(X)%*%X
# Estimates, compare with fm$coef
bHat2=solve(XtX,Xy)
bHat2

# To get SEs we need an estimate of the error variance
fitV = X%*%bHat2  ## Fitted values from ourmodel
eHat=y-fitV     ### Residual
plot(eHat)        ## Sequence plot of residual
plot(eHat ~ GOUT$su) ### ehat vs x
plot(eHat ~ fitV)   ### Residual vs fitted value plot
vEHat=sum(eHat^2)/(nrow(GOUT)-ncol(X)) # Sum of squares of errors divide by n-rank(X)
SE=sqrt(diag(solve(XtX)*vEHat)) ## SE of bHat
t_stat=bHat2/SE
## P-values (under normality first, and under t then...)
pvaluesN=pnorm(abs(t_stat),lower.tail=F)*2
pvaluesT=pt(abs(t_stat),df=length(y)-ncol(X),lower.tail=F)*2

summary(Model1_Gout)

cbind(bHat2,SE,t_stat,pvaluesT)
```

(a) Report your results and comment on any difference between this and results from 1(a)

The table is shown above in the matrix with the only difference being the intercept values. 

(b) Report the code you used to produce the results reported in 2(a)

```{r}
# Code is shown above. 
```

3. Using the estimate effects reported in 1(a) above; produce a plot with age in the x-axis, expected serum urate in the y-axis and lines for male/white, female/white, male/black, female/black. Use a range from 30 to 70 for the x-axis values. Clearly distinguish the different links (Use different colors on different line types). 

```{r}
#Base Plot
plot(x=GOUT$age ,y=GOUT$su, 
     xlab = "Age",
     ylab = "Serum Urate")
```

```{r}
## Prediction equations
bHat1=coef(Model1_Gout)
bHat1
plot(su~age,data =GOUT,
     xlab = "Age",
     ylab = "Serum Urate")

# Prediction Equation for Age: Base, no variables added
#Int= bHat1[1]
#Slope=bHat1[4]
#Age=seq(from=30,to=70,by=1)
#lines(x=Age,y=Int+Age*Slope,col=4)

# Prediction Equation for Age: Male, White
Int= bHat1[1]+bHat1[3]
Slope=bHat1[4]
Age=seq(from=30,to=70,by=1)
lines(x=Age,y=Int+Age*Slope,col="red",lwd=2,lty="dotdash")

# Prediction Equation for Age: Female, White
Int= bHat1[1]+bHat1[1]+bHat1[2]
Slope=bHat1[4]
Age=seq(from=30,to=70,by=1)
lines(x=Age,y=Int+Age*Slope,col="blue",lwd=2,lty="dotted")

# Prediction Equation for Age: Male, Black
Int= bHat1[1]
Slope=bHat1[4]
Age=seq(from=30,to=70,by=1)
lines(x=Age,y=Int+Age*Slope,col="black",lwd=2,lty="solid")

# Prediction Equation for Age: Female, Black 
Int= bHat1[1]+bHat1[2]
Slope=bHat1[4]
Age=seq(from=30,to=70,by=1)
lines(x=Age,y=Int+Age*Slope,col="green",lwd=2,lty="dashed")

legend(70, 12, legend=c("Male-White", "Female-White","Male-Black","Female-Black"),
       col=c("red", "blue","black","green"), lty=1:2, cex=0.8)

```

```{r}
Slope
```
Clearly interpret the slope for eaach line. 

The slope for all the lines is 0.02673734. What is changing is the intercept values.
